\chapter{Bacteriophages}
In this chapter we will provide more detailed description of bacteriophages, which will be the central topic of our work.
We will describe what they are and where we can find them.
We will also show currently valid taxonomical classification and the structure of a typical representant of this group of organisms.
Next we will explain their life cycle and we will point out some of their use.
% At the end of this chapter we will point out what was already done in this field and what we would like to achieve.

\section{Origin of bacteriophages}
Bacteriophages belong to the group of viruses, which have the capability to infect and replicate within bacterial hosts.
It is estimated that they are also the most abundant group of entities on the earth with their estimated count around $10^{31}$.
In comparison, it is estimated the count of all bacteria on planet Earth is around $10^{30}$.
They are also one of the most highly diverse group in the biosphere, with their genomes containing from a few genes to as many as hundreds of genes.
We can find them in every place on earth, where bacteria are able to live, even inside our bodies.
It is believed that one of the most saturated location of their occurrence is sea, with  $9\cdot 10^8$ virions per milliliter of seawater at the surface and 70\% of bacteria infected \cite{virioplankton}. % http://mmbr.asm.org/content/64/1/69.full
Thanks to the high abundance of bacteriophages, their impact on shaping of the environment is outstandingly significant.
% + circulating carbon?
% + they contribute to horizontal gene transfer via transduction and transformation

\section{Taxonomical classification}
% not sure if I want this section, but taxonomy of phages should probably be mentioned somewhere
% https://www.ncbi.nlm.nih.gov/pmc/articles/PMC523225/
Classification of bacteriophages is difficult mainly due to their high diversity and genome mosaicism. \cite{phagetax}


\section{Structure of a typical bacteriophage}
Given the high variety of bacteriophages, they come in a lot of different sizes and shapes.
Each bacteriophage consists of genetic information in form of DNA or RNA and capsid, a protein coat usually composited from higher number of the same protein units.
In this section we will describe the most typical form of bacteriophages, which is abundantly found in nature.
As we can see on the picture, this bacteriophages body consists of head and tail.
Both of these parts are created from proteins and in addition head contains DNA or RNA of the bacteriophage genome.
The tail is used to attach and to inject the genetic code to bacteria.
At the end of tail, there are proteins, which are able to bind to specific receptors on the surface of bacteria.
Thanks to this mechanism, bacteriophages tends to have high specificity in selection of their prey.

\section{Life cycle of bacteriophages}
Generally, there are 2 strategies, how bacteriophages try to secure their survival.
The first strategy is called lysogenic cycle and the second strategy is called lytic cycle.
Viruses tend to use these strategies in different proportions, but usually prefer to choose one of them.
The lysogenic cycle results in incorporations of phage genetic information into host DNA or creation of circular replicon in bacterial cytoplasm.
By this approach genetic material of phage inside host, called prophage, is duplicated together with the host genome and after cell division, both of daughter cells contain bacteriophages DNA.
In dependence on the following events, viral DNA can be released from hosts genome and start proliferation of new phages via the lytic cycle.
The lytic cycle is characterized by the lysis of bacterial cells membrane and their subsequent death.
It starts by injecting bacteriophage genome into bacteria.
After this step virus is not incorporated, but it compromises bacterial translation apparatus to produce more viruses.
Once enough virions have been produced, special viral proteins dissolve bacterial cell and virions are released into surrounding space.

\medskip
(picture of life cycle)
\medskip

\section{Potential usage of bacteriophages}
The ability to kill or alter behavior of bacteria with high specificity makes them valuable target for research.
Humankind tried to harness the power of phages since their discovery in year 1917.
This discovery is attributed to French-Canadian microbiologist Félix d'Hérelle, who experimented extensively with phages and introduced concept of phage therapy.
Unfortunately, after the discovery of antibiotics, research in this field suffered from insufficient funding and the development was significantly slowed down.
Nowadays, when humanity face the threat of multiresistant bacteria and phages could bring a new methods into the battle against pathogenic bacteria, the research in this field begin to flourish.
We can see the potential of their use in the field of food industry as the substances prolonging the life and improving the quality of our food,
in the field of pharmacy as medicines for bacterial infections and everywhere else, where we encounter the need to control the lives of microbial communities.

\section{Current state of knowledge}
\cite{phage}
